---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   #Pour avoir plusieurs lignes qui suivent. Commencer à 2 caractères
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal 00679993
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >  # Pour coordonne les packages LateX (supplémentaires)
  \usepackage{amsmath}
bibliography: references.bib # .bib à activer seulement pour git et word format (pandoc fait la biblio)
csl: 
lang: english # french
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook
#   bookdown::pdf_book:
#     base_format: EcoFoG::article
#     keep_tex: TRUE
#  bookdown::word_document2
#Mettre un code dans les figures pour indiquer le code d'où elle provient. ../ pour référer à des sous dossiers
---


<!-- Options de knitr et de R (librairies, ...) --> 

```{r Options, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=T, warning=FALSE, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40), out.width = '40%',fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package)}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
source("Scripts/DrawingTools.R")
#source("Scripts/GithubPages.R")
```

# Introduction

The large areas covered with tropical forests worldwide hold crucial economic, social and cultural value. 
They provide wood and multiple non-timber forest products, shelter a diversified fauna, regulate local climate, support carbon, water and nutrient cycles, and ensure cultural and human well-being. 
The simultaneous increase of forests products demand and substantial climatic changes heightened the pressure on the remaining forests [@Gibson2011a; @Morales-Hidalgo2015] and threaten the maintenance of communities structure, composition and functioning and the dynamics shaping them [@Anderson-Teixeira2013;@Sist2015].

In tropical forests, a constant range of disturbances determines the ecological rules that shape the communities, through the reallocation of resources (as light, heat and water fluxes), changes in interspecific relations, etc.
The response of communities to disturbance is then the cornerstone of tropical forests ecology, modelling and management [@White2001;@Chazdon2003a].
Disturbance recovery have been largely studied through structural parameters as aboveground biomass, tree height or stem density that all reflect rapid and measurable changes of ecosystems functioning asses crucial issues about the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016].
In this perspective tree species diversity is inevitable as it determines ecosystems productivity, stability and functioning through the interspecific complementarity that enhances resources use, nutrients storing, and hazards mitigation [@Tilman2014].
Reveal diversity dynamics is besides all the more important that disturbance, modifying the environment and interspecific relations, suerly impacts trees composition and diversity.

Until now, although, short-term dynamics demonstrated negligible or even positive impacts on communities diversity in accordance with the intermediate disturbance hypothesis (IDH) [@Molino2001;@Kariuki2006a;@Berry2008a] the long-term response of communities composition and diversity to disturbance remains unclear.
The IDH predicting a maximized species diversity at intermediate level of disturbance have besides mainly been tested through species richness, which gives limited or misleading information on forests recovery and functioning [@Martin2015;@Chaudhary2016].
Studies of communities response to disturbance should encompass taxonomic composition and evenness, which only can assess some conservation issues and the ecological rules that determine communities abundance distribution [@Magurran1988;@Lavorel2002;@Bellwood2006].
Beyond taxonomic metrics, communities recovery should also encompass functional diversity metrics [@Moretti2009;@Baraloto2012a] through functional traits, that are species biological attributes relating to individuals fitness and ecosystem properties [@Violle2007b;@Scheiter2013].
Functional traits were largely adopted and integrated within a workable framework to assess forests dynamics and functioning [@Diaz2005;@Villeger2008a] and a vast litterature revealed some major functional traits determinant for species ecology and performance.
These major traits proved relevant to assess the ecological rules shaping species assembly: post-disturbance trajectories in tropical rainforests, for example, are assumed to be driven by environmental filters fostering disturbance resistant species with rapid growth and efficient resources acquisition [@Molino2001;@Haddad2008].
Communities dynamics would then be reflected by a shift from "conservative" species displaying long-lived, dense tissues growing slowly but with scarce resources, to "acquisitive" species with fast-growing, light tissues allowing fast resources acquisition [@TerSteege2001;@Reich2014;@Herault2011].
In this perspective a combination of resource acquisition traits (leaf area,density and chlorophyll content and stem specific gravity and bark thickness) and life history traits of growth and reproduction (seed mass and maximum height) already proved relevant [@Wright2004;@Westoby2006a;@Chave2009b].

Here we investigated over 30 years the response of 75ha of forests plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem biomass removed.
We made use of a large functional traits database browsing major leaf, stem and seed traits and species maximum height to draw communities taxonomic and functional composition and diversity trajectories after disturbance.
Specifically, we (i) tested the validity of the IDH arguing for enhanced diversity after moderate disturbance and (ii) confirmed the changes in ecological rules shaping communities, from competitive exclusion between "conservative", slow-growing tree species to environmental filtering of "acquisitive", fasting-growing ones.
Eventually we examined the different trajectories to (iii) resolve the resilience of tropical forests and question their future dynamics regarding their state 30 years after disturbance.


# Material and Methods

### Study site 

Paracou station in French Guiana (5°18'N and 52°53'W) is located in a lowland tropical rain forest corresponding to a tropical wet climate with mean annual precipitation averaging 2980 mm.y^-1^ (30-y period) with a 3-months dry season (< 100 mm.months^-1^) from mid-August to mid-November, and a one-month dry season in March [@Wagner2011]. 
Elevation ranges between 5 and 50 m and mean annual temperature is 26°C. 
Soils correspond to thin acrisols over a layer of transformed saprolite with low permeability generating lateral drainage during heavy rains.
The experiment corresponds to a network of twelve 6.25ha plots that have undergone a gradient of three logging, thinning and fuelwood treatments \@ref(tab:Tab1).
Disturbance treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Blanc2009]. 

Table : (\#tab:Tab1) Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.

------------------------------------------------------------------------------------------------------------------------
                          ***Timber***              ***Thinning***          ***Fuelwood***           **%AGB lost**         

------------------  ------------------------ ------------------------- ------------------------ ------------------------
    ***Control***                                                                                          0            

         ***T1***    DBH ≥ 50 cm,                                                                                       
                     commercial species,                                                                [12%-33%]         
                     ~ 10 trees/ha                                                                                      

         ***T2***    DBH ≥ 50 cm,              DBH ≥ 40 cm,                                                                 
                     commercial species,       non-valuable species,                                    [33%-56%]         
                     ~ 10 trees/ha             ~ 30 trees/ha                                                              

         ***T3***    DBH ≥ 50 cm,              DBH ≥ 50 cm,               40 cm ≤ DBH ≤ 50 cm                               
                     commercial species,       non-valuable species,      non-valuable species,         [35%-56%]         
                     ~ 10 trees/ha             ~ 15 trees/ha              ~ 15 trees/ha                                     
------------------------------------------------------------------------------------------------------------------------

### Inventories protocol and dataset correction

The study site corresponds to a tropical rainforest with a dominance of Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical families. 
In the twelve experimental plots of the experiment, all trees above 10 cm DBH were mapped and measured annually since 1984. 
During inventories, trees were first identified with a vernacular name assigned by the field team, and afterward with a scientific name assigned by a botanist during regular botanical campaigns. 
In 1984, specific vernacular names were given to 62 commercial or common species whereas other less common species were identified under two identifiers only separating trees and palm trees. 
The botanical campaigns carried every 5 to 6 years to identify all trees at the species level only started in 2003 and identification practices varied among plots and successive campaigns. 
This raised methodological issues as vernacular names usually correspond to different botanical species, resulting in significant taxonomic uncertainties that had to be propagated to composition and diversity metrics throught a Bayesian framework. 
The uncertainty propagation was done by the replenishment of inventories completed at genus level from real incomplete ones on the basisof vernacular/botanical names association. 
Vernacular names were replaced through multinomial trials (M~v~([*s~1~, s~2~, …, s~N~*] ,[*α~1~, α~2~,…, α~N~*]) ) based on the observed association probability [*α~1~, α~2~,…, α~N~*] between each vernacular name *v* and the species [*s~1~, s~2~, …, s~N~*] recorded in the inventory. 
To avoid remaining identification caveats and consider complete inventories, the simulated inventories were then reported at genus.
See appendix 1 and @Aubry-Kientz2013 for the detailed methodology.

The functional diversity metrics used a dataset for 6 functional traits representing leaf economics (leaves thickness, toughness, total chlorophyll content and specific leaf area, the leaf area per unit dry mass) and wood economics (wood specific gravity and bark thickness), and life history traits (maximum specific height and seed mass).
Traits database came from the BRIDGE project (http://www.ecofog.gf/Bridge/index.html) where traits values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou. 
Missing trait values were filled using multivariate imputation by chained equation (mice) restricted to samples pertaining to the next higher taxonomic level, in order to account for the phylogenetic signal of the functional traits. 
The dataset comprised 294 botanical species pertaining to 157 botanical genus. 
Whenever a species inventoried was not in the dataset, it was attributed a set of traits values randomly sampled among species of the same next higher taxonomic level. 
All composition and diversity metrics corresponded to the expectancy obtained after 50 iterations of the taxonomic uncertainty propagation framework and of the filling process of missing trait values. 


### Composition and diversity metrics

The analysis of taxonomic and functional composition of plots along time was visualized in a two-dimensional ordination space based on non-metric dimensional scaling of succesive floristic or functional inventories. 
Plots trajectories along time was reported comparatively to the inventories in 1989, 5 years after disturbance, which corresponded to first inventory with a sufficient degree of uncertainty (<30% of undetermined trees). The inventories dissimilarity compared to the reference 1989 inventory was reported using occurrence-based (Jaccard) and abundance-based (Bray-Curtis) similarity measures. 
The trajectory of inventories along time was visualized with the euclidean distance in the two-dimensional ordination space to the 1989 inventory.
The functional trajectories were also visualized with the community weighted means (CWM) of the leaf and stem and data life traits. 
To determine whether communities’ assembly deviated from the expectation of a random process we compared the observed similarity patterns to those generated by 50 repetitions of stochastic null models. 
The null model for taxonomic composition randomly shuffled individuals among plots while preserving the overall species abundance and plots’ tree density. 
The null model for functional composition produced random traits assemblages by permuting the set of functional traits values among species. 

Plots taxonomic diversity corresponded to species Richness and the Hill number translation of Shannon and Simpson indices. 
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q). 
This order q grasps the balance between richness and evenness in the community as it determines the weight of common species in the diversity metric, with an growing emphasis on dominant species as q increases. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and data life traits.

# Results

Over time, 104423 trees and 696 botanical species were recorded. 
The disturbance proved to significantly impact for more than 30 years the functional and taxonomic composition and diversity of the stands. 
 
### Taxonomic and functional composition
 
*Composition trajectories* _ Trajectories of taxonomic and functional composition after disturbance were examined through the ordination of successive inventories from 1989 (5 years after disturbance) in a two-dimensional space based on NMDS analysis. 
Classifications were based on either abundance-based Bray-Curtis \@ref(fig:Fig1) or incidence-based Jaccard dissimilarity (data not shown).

Both taxonomic and functional composition were substantially affected by disturbance and positively correlated with its intensity.
*The distance from the initial condition was significantly time-dependent (P<0.01).* 
The taxonomic composition remained significantly different from the 1989 reference inventory until 30 years for all disturbed plots \@ref(fig:Fig1). 
From 20 years after disturbance, though, the distancing decreased which suggested a return toward the reference 1989 inventory and its resilience to disturbance \@ref(fig:Fig1),\@ref(fig:Fig2). 
The functional composition distancing from the reference inventory was also positively correlated with disturbance intensity and similarly stabilized or reduced (for 5 of T1, T2 and T3 plots) from 20 years after disturbance \@ref(fig:Fig2). 
*The coordinates of the functional traits were measured in the two-dimensional ordination space that mapped plots evolution along time (graphs not showed) which revealed that the trajectories of disturbed plots along time headed toward acquisitive functional strategies (from high SWG to high SLA and chlorophyll content).*

```{r Fig1, out.width = '100%', fig.asp = 1.2/2, echo=FALSE, fig.cap="Trajectories of the plots in terms of **(a)** flora composition and **(b)** functional composition regarding the 6 leaf and stem functional traits,the maximum allometric height and seed mass class in the two-dimensional space from the NMDS performed for the 30 years after disturbance. Distance matrix for NMDS were computed from the Bray-curtis dissimilarity between successive inventories. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3)."}
load("DB/FunctionalComposition_forGraph")
load("DB/TaxoComposition_forGraphs")

TaxoCompo(MatrepTaxo)
FunCompo(MatrepFun)

```



```{r Fig2, out.width = '100%', fig.asp = 1.2/2, echo=FALSE,fig.cap="Trajectories of the distance to initial condition of the 30 sampled years in the two-dimensional space from the NMDS of **(a)** taxonomic composition at genus level and **(b)** functional composition. Distance are abundance-based Bary-Curtis metric. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3). The 0.025 and 0.975 percentile correspond to the variance observed for 50 iteration of the taxonomic uncertainty propagation and functional trait filling processes. "}

TaxoDist(MatrepTaxo)
FunDist(MatrepFun)

```

*Traits community weighted means* _ The trajectories of the community weighted means (CWM) of the 8 functional traits were measured at plot level \@ref(fig:Fig3).

```{r Fig3, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the CWM until 30 years after disturbance of 4 leaf traits (Leaf thickness, chlorophyll content, toughness and specific area) and the 2 stem traits (Wood density and barck thickness). The CWM were measured at plot level and averaged among plots of the same treatment (green for control, blue for T1,orange for T2 and red for T3). The average CWM per treatment are drawn (solid line) drawn along with the 0.025 and 0.975 percentile (gray envelope) observed for 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes."}
load("DB/CWM")

CWMdraw(CWM)

```

```{r Fig4, echo=FALSE, out.width='100%', fig.asp=2/5,fig.cap="Trajectories of seed mass classes proportions until 30 years after disturbance. Plot treatments are indicated with line colors in green for control, blue for T1, orange for T2 and red for T3. The proportion reported are the median values (solid line) and the 0.025 and 0.975 percentile (gray envelope) observed for 50 iteration of the taxonomic uncertainty propagation. only species which seed mass values were reported are considered (*i.e* 414 out of 696 species recorded in total)."}
load("DB/Smass_ClassProp")

SeedMassProp(Smass)

```

### Disturbance impact on diversity

*Taxonomic diversity* _ Trajectories of Richness, Shannon and Simpson taxonomic diversity were examined in relation to the 1994 inventories \@ref(fig:Fig5). 
*For the three indices the OLS regression confirmed the time-dependency of the diversity variations (P<0.01) and a covariance analysis confirmed the significant effect of the disturbance treatment (P<0.01).*

```{r Fig5, out.width = '100%', fig.asp = 1.5/3, echo=FALSE,fig.cap="Trajectories of the difference to reference inventory, 10 years after disturbance, until 30 years after disturbance for plots **(a)** Richness, **(b)** Shannon and **(c)** Simpson diversity. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3). The average value of the 4 leaf traits (Leaf thickness, chlorophyll content, toughness and specific area) and the 2 stem traits (Wood density and barck thickness) are drawn along with the 0.025 and 0.975 percentile observed for 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes."}
load("DB/ReplacementTraj_ForGraphs")

TaxoTraj(CompleteTaxo)

```

The evolution along time of the difference of Richness, Shannon and Simpson diversity compared to reference inventories in 1989 remained even for 30 years after disturbance for the control plots whereas disturbed plots showed significant increase, positively correlated with the disturbance intensity \@ref(fig:Fig5). 
The richness stabilized for all disturbed plots from 20 years after disturbance \@ref(fig:Fig5). 
The Shannon and Simpson diversity also stabilized 20 years after disturbance for T1 and T2 plots \@ref(fig:Fig5) whereas for 2 out of 3 plots T3 the Shannon and Simpson diversity continued increasing until 30 years after disturbance, illustrating a significant increase in communities’ evenness.

*Functional  diversity* _ The evolution along time of the Rao functinal diversity compared to reference inventories in 1989 remained even all along the survey for the control plots. 
All disturbed plots however showed an increase of the functional diversity that was positively correlated to the disturbance intensity \@ref(fig:Fig6). 30 after disturbance the diversity of disturbed plots remained significantly higher than this of control plots, thus whenever the initial disturbance intensity.
The slope of the initial diversity increase and the maximum diversity were both positively correlated with the disturbance intensity. 
*The OLS regression confirmed the time-dependency of the diversity variations (P<0.01) and the covariance analysis confirmed the significant effect of the disturbance treatment (P<0.01)*.
The trajectories of plot 7 from the T1 disturbance treatment was discarded because of too high taxonomic uncertainty preventing correct analysis.  

```{r Fig6, fig.asp=1/1, echo=FALSE,fig.cap="Trajectories of the plots Rao functional diversity from10 to 30 years after disturbance. Line colors represent the disturbance treatment (green for control, blue for T1,orange for T2 and red for T3). The value of the 4 leaf traits (Leaf thickness, chlorophyll content, toughness and specific area) and the 2 stem traits (Wood density and barck thickness) are drawn along with the 0.025 and 0.975 percentile observed for 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes."}
load("DB/FunctionalTraj_ForGraphs")

FunTraj(CompleteFun)
```

# Discussion

Our study showed the significant impact of disturbance on tropical forests communities.
The subsequent diversity trajectories confirmed the intermediate disturbance hypothesis debated for tropical forests through their correlation with disturbance intensity.
Besides it revealed the contrasting response of taxonomic and functional characteristics, specifically the decoupling between communities taxonomic evenness and their functional diversity and dominant functional traits values.
The long-term disturbance trajectories observed highlighted the unachieved but consistent recovery of communities assembly for the lowest disturbance intensity but questioned it after higher disturbance.

In line with the "vegetation quantity effect" suggesting that communities functional characteristics depends on dominant species rather that subtle changes in biodiversity [@Grime1998]. This would result in a faster recovery of ecosystem processes rates than composition and diversity as demonstrated for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015;@Guariguata2001].  
This highlighted the functional overlap between species in tropical forests, typical of their huge biodiversity, and intricately questioned the different facets of communities resilience.
Ecosystem processes grasped through the set of functional traits retained here however do not assess communities resilience and future dynamics.

In line with the assumtions of the IDH the observed diversity trajectories confirmed the increase of communities evenness and functional richness after disturbance.
The consistency of the IDH has been debated for the higly diverse tropical forests, some observing the importance of environmental filters to shape communities assembly and others arguing for a genuine stochastic recovery [@Hubbell1999;@Molino2001].
Consistently to the trajectories drawn here, a previous study of the same experiment revealed the directional changes in communities composition reflecting environmental filtering after disturbance [@Molino2001].
Disturbance regimes brought a unimodal response depending on disturbance intensity as suggested by the outcome of 10 years and 20 years post-disturbance analysis[@Molino2001;@Baraloto2012a].
As observed after logging in a Boranean rainforest [@Cannon1998] or after cyclones in the Solomon islands [@Burslem2000], communities diversity showed a unimodal response to disturbance and a maintenance of diversity levels, except for one of the intensively disturbed community which evenness was still increasing \@ref(fig:Fig5). 
Despite the recovery of communities diversity, though, taxonomic composition was clearly altered \@ref(fig:Fig3) implying a long-term, if not permenant, composition change after disturbance as observed in East African tropical forests [@Bonnell2011].


Regarding the functionl trajectories and dominant trait values of disturbed communities \@ref(fig:Fig3), disturbance appeared to favor acquisitive strategies, *i.e* species with high SLA, low wood density, small seeds, etc [@Westoby1998;@Wright2004;@Reich2014].
The comparison of observed trajectories to a null model with random traits set shuffling showed an oriented functional shift after disturbance which suggest additional biotic and abiotic filters for the 10 to 15 years after disturbance.
This means a shift from exclusive competition favoring competitive species before disturbance to environmental filtering selecting acquisitive and disturbance resistant species.
In particular, the evolution seed mass classes proportions \@ref(fig:Fig4) revealed a clear increase in small seeded, therefore large dispersive, species [@TerSteege2001;@Flores2006;@Haddad2008].

The diversity and composition functional trajectories showed a decoupling between functional and taxonomic after the most intense disturbance (treatments T2 and T3).
Plots functional characteristic returned to their initial values, similar to those of control plots, faster than their taxonomic counterparts \@ref(fig:Fig2;fig:Fig3;fig:Fig4).
As functional traits are the most direct link between biodiversity and ecosystem functioning [@Diaz2005] the recovery of communities functional characteristics would go along the recovery of ecosystems processes although taxonomic composition remain altered [@Guariguata2001].
Such decoupling suggested the consistency of functional overlap and maybe functional versatility, implying a variety of functioning from the same characteristic, in the pre-disturbance condition.
Functional overlap has been observed in highly diverse ecosystems [@Bellwood2006] and suggested as one mechanisms for high diversity maintenance.
Knowing the decoupling between taxonomic and functional trajectories, this functional overlap may not be recovering after disturbance and therefore radically modify communities resilience and resistance [@Trenbath1999;@Elmqvist2003;@Diaz2005]. 
Given the impaired functional overlap, the persitant compositional changes regarding disturbance resistant species [@Haddad2008] and probably lianas or epiphytes [@Martin2013], and the modification of environmental conditions like soils nutrient cycling and compaction [@Olander2005] the entirety of communities recovery after intense disturbance is deeply questioned [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Although communities recovered the core of their initial functioning, composition seemed permanently altered after intense disturbance.
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk of cornerstone species loss and unexpected ecological consequences [@Diaz2005;@Gardner2007].

To confirm this long-term communities changes and highlight the ecological rules shaping communities assembly, we suggest a focus on recruitment dynamics, that would follow the diversity and functional characteristic of recruited communities and, by extension, of the forest to be.

-	Important differences among plots of same treatments: possible impact of initial diversity composition, local abiotic parameters (topography, soils, surrounding stands). Impacts depend on the disturbance intensity, but further parameters may be involved. 


`r if (knitr:::is_html_output()) '# References {-}'`

---
bibliography: ["references.bib"]
biblio-style: "apalike"
link-citations: true
---

